The role of aneuploidy in the evolution of cancer drug resistance

### 2023-06-14 EvolutionaryRescue1.tex
add todo

@@

We are interested in the mean conditional time, $\tau_1=\mathbb{E}\left[T_1 \mid T_1<\infty\right]$, which is given by
\begin{equation}\label{meantime1}
\begin{aligned}
\tau_1 =
\int_{0}^{\infty}{t f_1(t) \d t} = 
\frac{\int_{0}^{\infty}{tR(t)\exp\left(-\int_0^{t}{R(z)\d z}\right) \d t}}{p_{rescue}} = 
\frac{\int_{0}^{\infty}{\exp\left(-\int_0^{t}{R(z)\d z}\right) \d t}}{p_{rescue}}
\end{aligned}
\end{equation}
after applying integration by parts.
Therefore, plugging \cref{integralR,rescue_prob2} in \cref{meantime1}, 
\begin{align}\label{limitapprox}
\tau_1 =
\frac{\int_{0}^{\infty}{\exp\left(-u p_a N t\right) \d t}}{1-\e^{-Np_w}} \approx
\left(1+\e^{-Np_w}\right)\int_0^\infty\e^{-u p_a N t} \d t =
\frac{1+\e^{-Np_w}}{u p_a N} ,
\end{align}
where we use the approximation $\left(1-\e^{-Np_w}\right)^{-1}\approx1+\e^{-Np_w}$ and integrate the exponent.
\Cref{MeanTimeGrowthAneuploidyPlot} show the agreement between this approximating and simulation results for intermediate and large tumor sizes.

@@

We wish to obtain a simpler formula for $\tau_2$ in an analogous way to \eqref{limitapprox}. For this, we make use of the following expansions:
\begin{align*}
\frac{\e^{\Delta_w\tau}-1}{\Delta_w}&=\frac{1+\Delta_w\tau+\frac{\Delta_w^2\tau^2}{2}+O(\tau^3)-1}{\Delta_w}=\tau+\frac{\Delta_w}{2}\tau^2+O(\tau^3).\\
\frac{\e^{\Delta_a\tau}-1}{\Delta_a}&=\frac{1+\Delta_a\tau+\frac{\Delta_a^2\tau^2}{2}+O(\tau^3)-1}{\Delta_a}=\tau+\frac{\Delta_w}{2}\tau^2+O(\tau^3),
\end{align*}
which allow us to write:
\begin{equation*}
\frac{\e^{\Delta_w\tau}-1}{\Delta_w}-\frac{\e^{\Delta_a\tau}-1}{\Delta_a}\approx\frac{\left(\Delta_w-\Delta_a\right)\tau^2}{2}.
\end{equation*}
As a result, the integrand in \eqref{meantimet2} can be written as:
\begin{align*}
&\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_w\tau}-1}{\Delta_w}-\frac{\e^{\Delta_a\tau}-1}{\Delta_a}\right)-vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}\right]\approx\exp\left(-uvNp_m\tau^2-vNp_m\tau\right)\\
&=\exp\left(\frac{vNp_m}{2}\right)\exp\left[-\frac{uvNp_m}{2}\left(\tau+\frac{1}{u}\right)\right].
\end{align*}
Consequently, the mean time $\tau_2$ is obtained to be:
\begin{align}\label{limitapprox2}
\tau_2\approx\left[1+\exp\left(-Np_w\right)\right]\exp\left(\frac{vNp_m}{2u}\right)\frac{\erfc\left(\sqrt{\frac{vNp_m}{2u}}\right)}{\sqrt{\frac{2uvNp_m}{\pi}}},
\end{align}
where $\erfc$ is the complementary error function. We plot the expansion \eqref{limitapprox2} in Figure \ref{MeanTimeGrowthAneuploidyPlot} and observe that it is a very good fit for large values of the initial wildtype population size.

If we select only linear terms in the following expansions:
\begin{align*}
\frac{\e^{\Delta_w\tau}-1}{\Delta_w}&=\frac{1+\Delta_w\tau+O(\tau^2)-1}{\Delta_w}=\tau+O(\tau^2).\\
\frac{\e^{\Delta_a\tau}-1}{\Delta_a}&=\frac{1+\Delta_a\tau+O(\tau^2)-1}{\Delta_a}=\tau+O(\tau^2),
\end{align*}
we obtain the first order approximation for $\tau_2$:
\begin{align}\label{limitapprox3}
\tau_2\approx\left(1+\e^{-Np_w}\right)\int_0^\infty\e^{-uNp_m\tau}\,\d\tau=\frac{\left(1+\e^{-Np_w}\right)}{uNp_m},
\end{align}
which we plot in Figure \ref{MeanTimeGrowthAneuploidyPlot} and observe that it offer as a good a fit to \eqref{meantimet2} as \eqref{limitapprox2}. Additionally, we observe that for large initial wildtype populations sizes direct mutation drives evolutionary rescue while aneuploidy plays a role for intermediate sized tumors. This is consistent with the information obtained from \Cref{rescue_ratio} where aneuploidy improves the probability of evolutionary rescue only  for small and intermediate values of $N$.

>>>>

% TODO two papers that show aneuploidy provides resistance
% Lukow, Devon A., Erin L. Sausville, Pavit Suri, Narendra Kumar Chunduri, Angela Wieland, Justin Leu, Joan C. Smith, et al. 2021. “Chromosomal Instability Accelerates the Evolution of Resistance to Anti-Cancer Therapies.” Developmental Cell 56 (17): 2427-2439.e4. https://doi.org/10.1016/j.devcel.2021.07.009.
% Rutledge, Samuel D., Temple A. Douglas, Joshua M. Nicholson, Maria Vila-Casadesús, Courtney L. Kantzler, Darawalee Wangsa, Monika Barroso-Vilares, Shiv D. Kale, Elsa Logarinho, and Daniela Cimini. 2016. “Selective Advantage of Trisomic Human Cells Cultured in Non-Standard Conditions.” Scientific Reports 6 (June 2015): 1–12. https://doi.org/10.1038/srep22828.

@@

We are interested in the mean conditional time, $\tau_1=\mathbb{E}\left[T_1 \mid T_1<\infty\right]$, which is given by
\begin{equation}\label{meantime1}
\begin{aligned}
\tau_1 =
\int_{0}^{\infty}{t f_1(t) \d t} = 
\frac{\int_{0}^{\infty}{tR(t)\exp\left(-\int_0^{t}{R(z)\d z}\right) \d t}}{p_{rescue}} = 
\frac{\int_{0}^{\infty}{\exp\left(-\int_0^{t}{R(z)\d z}\right) \d t}}{p_{rescue}}
\end{aligned}
\end{equation}
after applying integration by parts.
Therefore, plugging \cref{integralR,rescue_prob2} in \cref{meantime1}, the probability density function of $T_1$ is given by:
\begin{align}
f_{T_1}\left(t_1\right)=R\left(t_1\right)\e^{-\int_0^{t_1} R(t)\d t}.
\end{align}
As a result, the time $T_1$ conditional on evolutionary rescue is given by:
\begin{align}
f_{T_1}\left(t_1|m_{t\rightarrow\infty}\neq0\right)=\frac{R\left(t_1\right)\e^{-\int_0^{t_1} R(t)\d t}}{1-\left(1-p_w\right)^N}.
\end{align}
The expectation of $T_1$ is:
\begin{align}\label{meantime1}
\tau_1=\mathbb{E}\left[T_1\right]&=\frac{\int_0^\infty\e^{-\int_0^\tau R(t)\d t}\d\tau}{1-\left(1-p_w\right)^N}=\frac{\int_0^\infty\e^{-uNp_a\frac{\e^{\Delta_w\tau}-1}{\Delta_w}}\,\d\tau}{1-\left(1-p_w\right)^N}.
\end{align}
The fration in the exponential of the integrand in \eqref{meantime1} can approximated as:
\begin{equation*}
\frac{\e^{\Delta_w\tau}-1}{\Delta_w}=\frac{1+\Delta_w\tau+O(\tau^2)-1}{\Delta_w}=\tau+O(\tau^2).
\end{equation*}
As a result, the mean time $\tau_1$ can be simplified:
\begin{align}\label{limitapprox}
\tau_1 =
\frac{\int_{0}^{\infty}{\exp\left(-u p_a N t\right) \d t}}{1-\e^{-Np_w}} \approx
\left(1+\e^{-Np_w}\right)\int_0^\infty\e^{-u p_a N t} \d t =
\frac{1+\e^{-Np_w}}{u p_a N} ,
\end{align}
where we use the approximation $\left(1-\e^{-Np_w}\right)^{-1}\approx1+\e^{-Np_w}$ and integrate the exponent.
\Cref{MeanTimeGrowthAneuploidyPlot} show the agreement between this approximating and simulation results for intermediate and large tumor sizes.

@@

We wish to obtain a simpler formula for $\tau_2$ in an analogous way to \eqref{limitapprox}. For this, we make use of the following expansions:
\begin{align*}
\frac{\e^{\Delta_w\tau}-1}{\Delta_w}&=\frac{1+\Delta_w\tau+\frac{\Delta_w^2\tau^2}{2}+O(\tau^3)-1}{\Delta_w}=\tau+\frac{\Delta_w}{2}\tau^2+O(\tau^3),\\
\frac{\e^{\Delta_a\tau}-1}{\Delta_a}&=\frac{1+\Delta_a\tau+\frac{\Delta_a^2\tau^2}{2}+O(\tau^3)-1}{\Delta_a}=\tau+\frac{\Delta_w}{2}\tau^2+O(\tau^3),
\end{align*}
which allow us to write:
\begin{equation*}
\frac{\e^{\Delta_w\tau}-1}{\Delta_w}-\frac{\e^{\Delta_a\tau}-1}{\Delta_a}\approx\frac{\left(\Delta_w-\Delta_a\right)\tau^2}{2}.
\end{equation*}
As a result, the integrand in \eqref{meantimet2} can be written as:
\begin{align*}
&\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_w\tau}-1}{\Delta_w}-\frac{\e^{\Delta_a\tau}-1}{\Delta_a}\right)-vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}\right]\approx\exp\left(-uvNp_m\tau^2-vNp_m\tau\right)\\
&=\exp\left(\frac{vNp_m}{2}\right)\exp\left[-\frac{uvNp_m}{2}\left(\tau+\frac{1}{u}\right)\right].
\end{align*}
Consequently, the mean time $\tau_2$ is obtained to be:
\begin{align}\label{limitapprox2}
\tau_2\approx\left[1+\exp\left(-Np_w\right)\right]\exp\left(\frac{vNp_m}{2u}\right)\frac{\erfc\left(\sqrt{\frac{vNp_m}{2u}}\right)}{\sqrt{\frac{2uvNp_m}{\pi}}},
\end{align}
where $\erfc$ is the complementary error function. We plot the expansion \eqref{limitapprox2} in Figure \ref{MeanTimeGrowthAneuploidyPlot} and observe that it is a very good fit for large values of the initial wildtype population size.

If we select only linear terms in the following expansions:
\begin{align*}
\frac{\e^{\Delta_w\tau}-1}{\Delta_w}&=\frac{1+\Delta_w\tau+O(\tau^2)-1}{\Delta_w}=\tau+O(\tau^2),\\
\frac{\e^{\Delta_a\tau}-1}{\Delta_a}&=\frac{1+\Delta_a\tau+O(\tau^2)-1}{\Delta_a}=\tau+O(\tau^2),
\end{align*}
we obtain the first order approximation for $\tau_2$:
\begin{align}\label{limitapprox3}
\tau_2\approx\left(1+\e^{-Np_w}\right)\int_0^\infty\e^{-uNp_m\tau}\,\d\tau=\frac{\left(1+\e^{-Np_w}\right)}{uNp_m},
\end{align}
which we plot in Figure \ref{MeanTimeGrowthAneuploidyPlot} and observe that it offer as a good a fit to \eqref{meantimet2} as \eqref{limitapprox2}. Additionally, we observe that for large initial wildtype populations sizes direct mutation drives evolutionary rescue while aneuploidy plays a role for intermediate sized tumors. This is consistent with the information obtained from \Cref{rescue_ratio} where aneuploidy improves the probability of evolutionary rescue only  for small and intermediate values of $N$.
